A weighted sampling algorithm for the design of RNA sequences with targeted secondary structure and nucleotides distribution

### 2013-03-08 ISMB/methods_ISMB.tex
Update

A targeted secondary structure $\Target$ is given as a non-crossing arc-annotated sequence,  where 
$\Target_i$ stands for the base-pairing position of position $i$ in $\Target$ if any (and, reciprocally, $\Target_{\Target_i}=i$), or $-1$ otherwise. 
In addition, we denote by $\gc(s)$ the number of occurrences of \Gb and \Cb in an RNA sequence $s$.

\subsubsection{Simplified energy model}
We use a simplified free-energy model which only includes additive contributions from stacking base-pairs. Using individual values from the Turner 2004 model (retrieved from the NNDB~\cite{Turner2010}). Given a candidate sequence $s$ for a secondary structure $S$, the free-energy of $S$ on $s$ is given by
\begin{align*}
  \ES(s,S) = \sum_{\substack{(i,j)\to (i',j')\in S\\ \text{stacking pairs}}}\ES^{\beta}_{s_is_j\to s_{i'}s_{j'}} 
\end{align*}
where $\ES^{\beta}_{ab\to a'b'}$ is set to $0$ if $ab=\varnothing$ (no base-pair to stack onto), the tabulated free-energy of stacking pairs $(ab)/(a'b')$ in the Turner model if available, or $\beta\in[0,\infty]$ for non-Watson-Crick/Wobble pairs (i.e. not in $\{\Gb\Ub,\Ub\Gb,\Cb\Gb,\Gb\Cb, \Ab\Ub\text{ or }\Ub\Ab\}$). This latter parameter allows one to choose whether to simply penalize invalid base pairs ($\beta>0$), or forbid them altogether ($\beta = +\infty$). 
Position-specific sequence constraints can also be enforced at this level (details omitted for the sake of clarity) by assigning to $\ES$ a $+\infty$ penalty (leading to a null probability) to the presence, at a given position, of a base that is incompatible with a user-specified constraint mask.

@@

In order to counterbalance the documented tendency of sampling methods to generate \Gb\Cb-rich sequences~\cite{Levin:2012kx}, we introduce a parameter $x\in\mathbb{R}^+$, whose value will influence the \GCContent of generated sequences. Let $\Target$ be a targeted secondary structure, then the pseudo-Boltzmann factor of a sequence $s$ is $\B_{x}(s,S)$  such that
\begin{equation}
\B_{x, \Target}(s) = e^{\frac{-\ES(s,\Target)}{RT}}\cdot x^{\gc(s)}
\label{def:genBoltz}
\end{equation}
where $R$ is the Boltzmann constant and $T$ the temperature in Kelvin.

Summing the pseudo-Boltzmann factor over all possible sequences of a given length $|\Target|$, one obtains the pseudo-partition function $\mathcal{Z}_{x,\Target}$, from which one defines the pseudo-Boltzmann probability $\Prob_{x,\Target}(s)$ of each sequence $s$, respectively such that 
\begin{align}\mathcal{Z}_{x,\Target} &= \sum_{\substack{|s|=|\Target|}}\B_{x,\Target}(s)& \text{and}&&
\Prob_{x,\Target}(s) &= \frac{\B_{x,\Target}(s)}{\mathcal{Z}_{x,\Target}}.\label{def:distribution}\end{align}

@@

Firstly, a dynamic programming algorithm computes $\mathcal{Z}\substack{(i,j)\\ [a,b]}$ the pseudo-partition function restricted to an interval $[i,j]$, assuming its (previously chosen) flanking nucleotides are $a$ and $b$ respectively. 
Remark that the empty interval only supports the empty structure, having energy $0$, so one has
\begin{equation}
	\forall i \in [0,n-1]:\, \Z{i+1,i}{a,b}=1.
	\label{eq:Z_in}
\end{equation}

@@

Despite its apparent quadratic complexity, the recurrence described in Subsection~\ref{sec:pf} can be computed in linear $\Theta(|\Target|)$ time and space, owing to the fact that, on any recursive call, the position $k$ is entirely determined by $i$.

\begin{theorem}Starting from the complete region $[0,n-1]$, Equation~\eqref{eq:Z_rec} can be computed in $\Theta(|\Target|)$ time and space complexity.
\end{theorem}
\begin{proof}

Consider the {\bf covering base-pair} $\CNBP{i}$ of a position $i$, defined as the base-pair $(i',j')$ such that $i'<i<j'$ and which minimizes $i-i'$. For the sake of simplicity, and without loss of generality, let us assume that there exists a virtual base-pair $(-1,n)$ in $\Target$, so that any position in $[0,n-1]$ has a covering base-pair.
Then, it is easily shown by induction that, starting from the complete interval $[0,n-1]$, the recursive calls only evaluate $\mathcal{Z}$ over intervals $[i,j]$ such that $j=j'-1$, $\CNBP{i}=[i',j']$. 
\CFramed{This property clearly holds initially for the complete interval $[0,n-1]$, owing to the virtual base-pair $(-1,n)$. Furthermore, if it holds for a given $(i,j)$, $\CNBP{i}=[i',j+1]$, then:
\begin{itemize} 
  \item If $i$ is unpaired (Case 1), then $\CNBP{i+1}=\CNBP{i}=[i',j+1]$, and the property is satisfied by the subsequent call over $[i+1,j]$. 
  \item If $i$ is paired with $j$ and stacking onto $(i-1,j+1)$ (Case 2), then one has $\CNBP{i+1}=[i,j]$ and the  subsequent call over $[i+1,j-1]$ satisfies the property.
  \item If $i$ is paired with some $k$ (Case 3), then $\CNBP{i+1}=[i,k]$ and $\CNBP{k+1}=[i',j+1]$. Therefore the subsequent calls, respectively over $[i+1,k-1]$ and $[k+1,j]$, both satisfy the property.
\end{itemize}
The case of an empty sequence does not cause a recursive call and must not be verified.
}
Since, in any interval $[i,j]$ satisfying the property, the ending position $j$ is completely determined by $i$, then there exists at most $\Theta(|\Target|)$ intervals that need to be evaluated, and we conclude on the overall complexity with the remark that $a$, $b$, $a'$ and $b'$ vary over bounded ranges and only contribute to a constant factor in the complexity.
\end{proof}

\begin{figure}
\resizebox{\textwidth}{!}{\input{FigStochasticBacktrack.pgf}}
\caption{Stochastic backtrack procedure over an interval $[i,j]$: Either position $i$ is left unpaired (top), a base-pair $(i,j)$ is formed, stacking onto $(i+1,j-1)$ (middle), or $i$ is paired somewhere else, defining two regions on which subsequent recursive calls are needed (bottom). Positions indicated in red are assigned during this stage of the backtrack.\label{fig:stochastic}}
\end{figure}
\subsubsection{Stochastic backtrack}
Once the pseudo-partition functions have been memorized, a stochastic backtrack starts from the complete interval $[0,|\Target|-1]$ and, at each step, chooses a suitable assignment for one or several positions, using probabilities derived from the precomputation, as illustrated by Figure~\ref{fig:stochastic}. One or several recursive calls over the remaining regions are then performed, as described in Algorithm~\ref{alg:back}. On each recursive call, the algorithm assigns at least one nucleotide to a -- previously unassigned -- position. Moreover, the number of executions of each loops is bounded by a constant. Consequently, the complexity of Algorithm~\ref{alg:back} is in $\Theta(|\Target|)$ time and space.

@@

Regarding the complexity, it was previously established~\cite{Waldispuhl2011} that, for each value of $x$, there exists constants $\mu_x$ and $\sigma_x$ such that the distribution of \GCContent asymptotically converges towards a normal law having expectation in $\mu\cdot n\cdot(1+o(1))$ and standard deviation in $\sigma\cdot\sqrt{n}\cdot(1+o(1))$.
Furthermore, the concentration of the distribution, asserted by its limited standard deviation, implies a growth in $\Theta(\sqrt{n})$ (resp. $\Theta(1)$, i.e. constant) for the expected number of attempts required to generate a valid sequence when $\kappa=0$ (resp. $\kappa\in\Omega(1/\sqrt n)$).

>>>>

A targeted secondary structure $\Target$ of length $n$ is given as a non-crossing arc-annotated sequence,  where 
$\Target_i$ stands for the base-pairing position of position $i$ in $\Target$ if any (and, reciprocally, $\Target_{\Target_i}=i$), or $-1$ otherwise. 
In addition, let us denote by $\gc(s)$ the number of occurrences of \Gb and \Cb in an RNA sequence $s$.

\subsubsection{Simplified energy model}
We use a simplified free-energy model which only includes additive contributions from stacking base-pairs. Using individual values from the Turner 2004 model (retrieved from the NNDB~\cite{Turner2010}). Given a candidate sequence $s$ for a secondary structure $\Struct$, the free-energy of any sequence $s$ of length $|\Struct|$  is given by
\begin{align*}
  \ES(s,\Struct) = \sum_{\substack{(i,j)\to (i',j')\in \Struct\\ \text{stacking pairs}}}\ES^{\beta}_{s_is_j\to s_{i'}s_{j'}} 
\end{align*}
where $\ES^{\beta}_{ab\to a'b'}$ is set to $0$ if $ab=\varnothing$ (no base-pair to stack onto), the tabulated free-energy of stacking pairs $(ab)/(a'b')$ in the Turner model if available, or $\beta\in[0,\infty]$ for non-Watson-Crick/Wobble pairs (i.e. not in $\{\Gb\Ub,\Ub\Gb,\Cb\Gb,\Gb\Cb, \Ab\Ub\text{ or }\Ub\Ab\}$). This latter parameter allows one to choose whether to simply penalize invalid base pairs ($\beta>0$), or forbid them altogether ($\beta = +\infty$). 
Position-specific sequence constraints can also be enforced at this level (details omitted for the sake of clarity) by assigning to $\ES$ a $+\infty$ penalty (leading to a null probability) in the presence of a base incompatible with a user-specified constraint mask.

@@

In order to counterbalance the documented tendency of sampling methods to generate \Gb\Cb-rich sequences~\cite{Levin:2012kx}, we introduce a parameter $x\in\mathbb{R}^+$, whose value will influence the \GCContent of generated sequences. For any secondary structure $\Struct$, the pseudo-Boltzmann factor of a sequence $s$ is $\B^{[x]}_{\Struct}(s)$  such that
\begin{equation}
\B_{\Target}^{[x]}(s) = e^{\frac{-\ES(s,\Target)}{RT}}\cdot x^{\gc(s)}
\label{def:genBoltz}
\end{equation}
where $R$ is the Boltzmann constant and $T$ the temperature in Kelvin.

Summing the pseudo-Boltzmann factor over all possible sequences of a given length $|\Target|$, one obtains the pseudo-partition function $\mathcal{Z}_{x,\Target}$, from which one defines the pseudo-Boltzmann probability $\Prob_{x,\Target}(s)$ of each sequence $s$, respectively such that 
\begin{align}\mathcal{Z}_{x,\Target} &= \sum_{\substack{|s|=n}}\B_{x,\Target}(s)& \text{and}&&
\Prob_{x,\Target}(s) &= \frac{\B_{x,\Target}(s)}{\mathcal{Z}_{x,\Target}}.\label{def:distribution}\end{align}

@@

Firstly, a dynamic programming algorithm computes $\Z{i,j}{a,b}$ the pseudo-partition function restricted to an interval $[i,j]$, assuming its (previously chosen) flanking nucleotides are $a$ and $b$ respectively. 
Remark that the empty interval only supports the empty structure, having energy $0$, so one has
\begin{equation}
	\forall i \in [0,n-1]:\, \Z{i+1,i}{a,b}=1.
	\label{eq:Z_in}
\end{equation}

@@

Despite its apparent quadratic complexity, the recurrence described in Subsection~\ref{sec:pf} can be computed in linear $\Theta(n)$ time and space, owing to the fact that, on any recursive call, the position $k$ is entirely determined by $i$.

\begin{theorem}Starting from the complete region $[0,n-1]$, Equation~\eqref{eq:Z_rec} can be computed in $\Theta(n)$ time and space complexity.
\end{theorem}
\begin{proof}

Consider the {\bf covering base-pair} $\CNBP{i}$ of a position $i$, defined as the base-pair $(i',j')$ such that $i'<i<j'$ and which minimizes $i-i'$. For the sake of simplicity, and without loss of generality, let us assume that there exists a virtual base-pair $(-1,n)$ in $\Target$, so that any position in $[0,n-1]$ has a covering base-pair.
Then, it is easily shown by induction that, starting from the complete interval $[0,n-1]$, the recursive calls only evaluate $\mathcal{Z}$ over intervals $[i,j]$ such that $j=j'-1$, $\CNBP{i}=[i',j']$. 
\CFramed{This property clearly holds initially for the complete interval $[0,n-1]$, owing to the virtual base-pair $(-1,n)$. Furthermore, if it holds for a given $(i,j)$, $\CNBP{i}=[i',j+1]$, then:
\begin{itemize} 
  \item If $i$ is unpaired (Case 1), then $\CNBP{i+1}=\CNBP{i}=[i',j+1]$, and the property is satisfied by the subsequent call over $[i+1,j]$. 
  \item If $i$ is paired with $j$ and stacking onto $(i-1,j+1)$ (Case 2), then one has $\CNBP{i+1}=[i,j]$ and the  subsequent call over $[i+1,j-1]$ satisfies the property.
  \item If $i$ is paired with some $k$ (Case 3), then $\CNBP{i+1}=[i,k]$ and $\CNBP{k+1}=[i',j+1]$. Therefore the subsequent calls, respectively over $[i+1,k-1]$ and $[k+1,j]$, both satisfy the property.
\end{itemize}
The case of an empty sequence does not cause a recursive call and must not be verified.
}
Since, in any interval $[i,j]$ satisfying the property, the ending position $j$ is completely determined by $i$, then there exists at most $\Theta(n)$ intervals that need to be evaluated, and we conclude on the overall complexity with the remark that $a$, $b$, $a'$ and $b'$ vary over bounded ranges, only contributing to a constant factor in the complexity.
\end{proof}

\begin{figure}
\resizebox{\textwidth}{!}{\input{FigStochasticBacktrack.pgf}}
\caption{Stochastic backtrack procedure over an interval $[i,j]$: Either position $i$ is left unpaired (top), a base-pair $(i,j)$ is formed, stacking onto $(i+1,j-1)$ (middle), or $i$ is paired somewhere else, defining two regions on which subsequent recursive calls are needed (bottom). Positions indicated in red are assigned during this stage of the backtrack.\label{fig:stochastic}}
\end{figure}
\subsubsection{Stochastic backtrack}
Once the pseudo-partition functions have been memorized, a stochastic backtrack starts from the complete interval $[0,n-1]$ and, at each step, chooses a suitable assignment for one or several positions, using probabilities derived from the precomputation, as illustrated by Figure~\ref{fig:stochastic}. One or several recursive calls over the remaining regions are then performed, as described in Algorithm~\ref{alg:back}. On each recursive call, the algorithm assigns at least one nucleotide to a -- previously unassigned -- position. Moreover, the number of executions of each loops is bounded by a constant. Consequently, the complexity of Algorithm~\ref{alg:back} is in $\Theta(n)$ time and space.

@@

Regarding the overall complexity of the algorithm, it was previously established~\cite{Waldispuhl2011} that, for each value of $x$, there exists constants $\mu_x$ and $\sigma_x$ such that the distribution of \GCContent asymptotically converges towards a normal law having expectation in $\mu_x\cdot n\cdot(1+o(1))$ and standard deviation in $\sigma_x\cdot\sqrt{n}\cdot(1+o(1))$.
Furthermore, the concentration of the distribution, asserted by its limited standard deviation, therefore the expected number of attempts required to generate a valid sequence when $\kappa=0$ (resp. $\kappa\in\Omega(1/\sqrt n)$) grows linke $\Theta(\sqrt{n})$ (resp. $\Theta(1)$, i.e. a constant), leading to the announced complexities.
